Convergence rate of MAP estimates \\ for the exponential family

### 2021-09-27 main.tex
Add proof of convergence for mirror descent

\section{Expected Relative Smoothness}
To avoid gradient variance assumption, we can take inspiration from expected smoothness, which is a consequence of convexity and smoothness, and which yields a general analysis of SGD. Can we get something similar with relativity and SMD? 
The analog of gradient norm is the Bregman between iterates, similar to the Homeomorphic EM rate.

\subsection{Convergence of Mirror Descent}
To get the descent lemma, apply successively relative smoothness, 3 points lemma, and relative strong convexity on $f$. Note that the strong convexity constant $m$ may be zero. Recall that $\ell_f(\natp, \natp^{t}) = f(\natp^t) + \langle \nabla f (\natp^t) ; \natp - \natp^t \rangle $ is the linearization of $f$ at $\natp^t$ evaluated in $\natp$. 
\begin{align}
	f(\natp^{t+1}) 
	&\leq \ell_f(\natp^{t+1}, \natp^{t})  + L \bregman(\natp^{t+1}; \natp^{t})
	& \text{(Relative $L$-smoothness)} \\
	& \leq \ell_f(\natp^{*}, \natp^{t})  + L \bregman(\natp^{*}; \natp^{t}) -L \bregman(\natp^{*}; \natp^{t+1}) 
	& \text{(3 Points Lemma w. $x=\natp^*$)} \\
	& \leq f(\natp^*) + (L - m) \bregman(\natp^{*}; \natp^{t}) -L \bregman(\natp^{*}; \natp^{t+1}) 
	& \text{(Relative $m$-strong convexity)}
\end{align}
Applying the 3 points lemma with \TODO{f}

@@

\subsection{Full Gaussian $\cN(m, \sigma^2)$}
Take the sufficient statistic $T(X) = (X, X^2)$. Then the mean parameter is $\mu = \E[T(X)] = ( m , m^2 + \sigma^2) = (\mu_1, \mu_2) $.
The log-density is written
\begin{align}
	\log p(X) 
	&= -\frac{(X-m)^2}{2 \sigma^2} - \half \log \sigma^2 + \cst \\
	&=  \frac{m}{\sigma^2} X -\inv{2 \sigma^2} X^2 - \frac{m^2}{2 \sigma^2} - \half \log \sigma^2 \\
	&=  \natp_1 X + \natp_2 X^2  - \logpart(\natp) \; .
\end{align}
From this formula, we can identify the duality map
\begin{align}
	\natp_1 &= \frac{m}{\sigma^2} = \frac{\mu_1}{\mu_2 - \mu_1^2} \\
	\natp_2 &= -\inv{2 \sigma^2} =  \frac{ - 1}{2(\mu_2 - \mu_1^2)}  < 0
\end{align}
and we can also explicit the log-partition as a function of the natural parameters
\begin{align}
	\logpart(m, \sigma) &= \frac{m^2}{2 \sigma^2} + \half \log \sigma^2 \\
	\logpart(\natp) &= \frac{\natp_1^2}{-4\natp_2} - \half \log(-\natp_2) \; .
\end{align}
We can perform a sanity check by deriving $\logpart$ and verifying that its gradient gives $\mu(\natp)$
\begin{align}
	\mu_1 &= \frac{\natp_1}{-2\natp_2}  \\
	\mu_2 & = \left( \frac{\natp_1}{-2\natp_2} \right)^2 + \inv{-2 \natp_2} \; .
\end{align}
Now let us discover the entropy
\begin{align}
	\conj (\mu) &= \mu_1 \natp_1 + \mu_2 \natp_2  - \logpart(\natp(\mu)) \\
	& = \frac{\mu_1^2}{\mu_2 - \mu_1^2}  - \frac{\mu_2}{2(\mu_2 - \mu_1^2)} - \frac{\mu_1^2}{2(\mu_2 - \mu_1^2)}  - \half \log(\mu_2 - \mu_1^2) \\
	&= \half \frac{\mu_1^2 - \mu_2}{\mu_2 - \mu_1^2} - \half \log(\mu_2 - \mu_1^2) \\
	& = - \half \log(\mu_2 - \mu_1^2) + \cst \; .
\end{align}
In this case, the Bregman divergences induced by the log-partition or the entropy are ugly, so we will not focus on them.

@@

>>>>

\section{Optimization Ideas}

@@

\subsection{Full Gaussian}
To model $\cN(m,\sigma^2)$, we take the sufficient statistic $T(X) = (X, X^2)$. Then the mean parameter is $\mu = \E[T(X)] = ( m , m^2 + \sigma^2) = (\mu_1, \mu_2) $.
The log-density is written
\begin{align}
	\log p(X) 
	&= -\frac{(X-m)^2}{2 \sigma^2} - \half \log \sigma^2 + \cst \\
	&=  \frac{m}{\sigma^2} X -\inv{2 \sigma^2} X^2 - \frac{m^2}{2 \sigma^2} - \half \log \sigma^2 \\
	&=  \natp_1 X + \natp_2 X^2  - \logpart(\natp) \; .
\end{align}
From this formula, we can identify the duality map
\begin{align}
	\natp_1 &= \frac{m}{\sigma^2} = \frac{\mu_1}{\mu_2 - \mu_1^2} \\
	\natp_2 &= -\inv{2 \sigma^2} =  \frac{ - 1}{2(\mu_2 - \mu_1^2)}  < 0
\end{align}
and we can also explicit the log-partition as a function of the natural parameters
\begin{align}
	\logpart(m, \sigma) &= \frac{m^2}{2 \sigma^2} + \half \log \sigma^2 \\
	\logpart(\natp) &= \frac{\natp_1^2}{-4\natp_2} - \half \log(-\natp_2) \; .
\end{align}
We can perform a sanity check by deriving $\logpart$ and verifying that its gradient gives $\mu(\natp)$
\begin{align}
	\mu_1 &= \frac{\natp_1}{-2\natp_2}  \\
	\mu_2 & = \left( \frac{\natp_1}{-2\natp_2} \right)^2 + \inv{-2 \natp_2} \; .
\end{align}
Now let us discover the entropy
\begin{align}
	\conj (\mu) &= \mu_1 \natp_1 + \mu_2 \natp_2  - \logpart(\natp(\mu)) \\
	& = \frac{\mu_1^2}{\mu_2 - \mu_1^2}  - \frac{\mu_2}{2(\mu_2 - \mu_1^2)} - \frac{\mu_1^2}{2(\mu_2 - \mu_1^2)}  - \half \log(\mu_2 - \mu_1^2) \\
	&= \half \frac{\mu_1^2 - \mu_2}{\mu_2 - \mu_1^2} - \half \log(\mu_2 - \mu_1^2) \\
	& = - \half \log (\mu_2 - \mu_1^2) + \cst \; .
\end{align}
In this case, the Bregman divergences induced by the log-partition or the entropy are ugly, so we will not focus on them.

@@

\subsection{Convergence of Mirror Descent}
\paragraph{Descent Lemma.}
To get the descent lemma, apply successively relative smoothness, 3 points lemma, and relative strong convexity on $f$. Note that the strong convexity constant $m$ may be zero. Recall that $\ell_f(\natp, \natp^{t}) = f(\natp^t) + \langle \nabla f (\natp^t) ; \natp - \natp^t \rangle $ is the linearization of $f$ at $\natp^t$ evaluated in $\natp$. 
\begin{align}
	f(\natp^{t+1}) 
	&\leq \ell_f(\natp^{t+1}, \natp^{t})  + L \bregman(\natp^{t+1}; \natp^{t})
	& \text{(Relative $L$-smoothness)} \\
	& \leq \ell_f(\natp^{*}, \natp^{t})  + L \bregman(\natp^{*}; \natp^{t}) -L \bregman(\natp^{*}; \natp^{t+1}) 
	& \text{(3 Points Lemma w. $x=\natp^*$)} \\
	& \leq f(\natp^*) + (L - m) \bregman(\natp^{*}; \natp^{t}) -L \bregman(\natp^{*}; \natp^{t+1}) 
	& \text{(Relative $m$-strong convexity)}
\end{align}

\paragraph{Monotonicity.}
Applying the 3 points lemma with $x=\natp_t$, 
\begin{align}
	f(\natp^{t+1}) \leq  f(\natp^t) -L (\bregman(\natp^{t}, \natp^{t+1}) + \bregman(\natp^{t+1} , \natp^{t} ))
\end{align}
shows the function value is monotonically decreasing, and we the gap at each step is greater than the symmetrized bregman between iterates.

\paragraph{No Strong-Convexity.}
The simpler case is when $m=0$, then we get a telescopic sum
\begin{align}
	\sum_{t=0}^{T-1} (f(\natp^{t+1}) - f(\natp^*)) 
	&\leq L \sum_t \bregman(\natp^{*}; \natp^{t}) - \bregman(\natp^{*}); \natp^{t+1}) \\
	&= L \bregman(\natp^{*}; \natp^{0}) - L\bregman(\natp^{*}); \natp^{T})\\
	&\leq L \bregman(\natp^{*}; \natp^{0}) \; ,
\end{align}
from which we can conclude thanks to the monotonicity of $f(\natp^t)$
\begin{align}
	f(\natp^{T}) - f(\natp^*)
	\leq  \frac{L}{T} \bregman(\natp^{*}; \natp^{0}) \; ,
\end{align}

\paragraph{Strong Convexity.} 
When $m>0$, we need to do some cooking with a coefficient $\frac{L-m}{L}$ to create the telescopic sum
\begin{align}
	\sum_{t=0}^{T-1} \left( \frac{L-m}{L} \right)^{T-1-t} (f(\natp^{t+1}) - f(\natp^*)) 
	&\leq \sum_{t=0}^{T-1} \frac{( L-m)^{T-t}}{L^{T-t -1}} \bregman(\natp^{*}; \natp^{t}) 
	- \frac{( L-m)^{T-t-1}}{L^{T-t-2}} \bregman(\natp^{*}); \natp^{t+1}) \\
	&= \frac{( L-m)^{T}}{L^{T-1}}  \bregman(\natp^{*}; \natp^{0}) 
	- L \bregman(\natp^{*}); \natp^{T}) \\
	&\leq L \left( \frac{L-m}{L} \right)^T \bregman(\natp^{*}; \natp^{0}) \; .
\end{align}
Then using the monotonicity of $f(\natp^t)$, we end up with the geometric sum of $\frac{L-m}{L} < 1$
\begin{align}
		\sum_{t=0}^{T-1} \left( \frac{L-m}{L} \right)^{T-1-t} (f(\natp^{t+1}) - f(\natp^*)) 
		&\geq (f(\natp^{T}) - f(\natp^*)) \sum_{t=0}^{T-1} \left( \frac{L-m}{L} \right)^{T-1-t} \\
		&\geq (f(\natp^{T}) - f(\natp^*)) \inv{1 - \frac{L-m}{L}} \\
		&= (f(\natp^{T}) - f(\natp^*)) \frac{L}{m} \; .
\end{align}
Finally we end up with the final iterate linear convergence rate
\begin{align}
	f(\natp^{T}) - f(\natp^*)
	\leq m  \left( 1 - \frac{m}{L} \right)^T \bregman(\natp^{*}; \natp^{0}) 
\end{align}
featuring $\frac{L}{m} $ the relative condition number.
